\large End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK

### 2020-05-27 doc/2020-SEFM/paper.tex
More details on SPARK in the Introduction

Several tools for GR(1) synthesis are available. 
For example, RATSY %(Requirements Analysis Tool with Synthesis) 
\cite{bloem2010ratsy} has a focus on circuit design and synthesizes designs 
encoded in BLIF, %(Berkeley Logic Interchange Format),
Verilog, and HIF. %(HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP 
%(Linear Temporal Logic MissiOn Planning)  
\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
TuLiP 
%(Temporal Logic Planning) 
\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
Slugs 
%(SmalL bUt Complete GROne Synthesizer) 
\cite{Ehlers2016} is architected to allow users to tailor synthesis algorithms,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
 %% -- TODO: put more details about input types that Salty allows? About what optimizations it does?
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable software implementations in a variety of programming languages, 
including Python, C++, Java, and now SPARK.

Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,'' 
tool implementation errors could still result in errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers. % with respect to their original GR(1) specifications. 
Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK. 
SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
SPARK translates code to the Why language for processing by Why3, 
and Why3 uses other automated provers -- including Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers by
annotating the code, e.g. with assertions and loop invariants. 
However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their GR(1) specifications in SPARK. 
To date, we have achieved this goal for a subset of properties comprising GR(1) specifications, 
i.e. system initial and transition properties, for moderately-sized controllers.

@@

We have extended the open source tool Salty\footnote{https://github.com/GaloisInc/salty/} to produce software implementations of controllers synthesized from GR(1) specifications in SPARK. 
Given a GR(1) specification in Salty format, Salty does some preprocessing to sanity check, optimize, and translate the specification to Slugs format. 
Salty then calls Slugs to synthesize a controller design from the processed specification. 
Slugs provides the option of returning the controller design as a Moore machine expressed in text format, 
which is what Salty uses to create an executable software implementation of the controller. 
This Moore machine encodes all the states of the controller, including a single initial state. 
It also encodes the transition relation between states, i.e. the controller's next state given the current state 
and the next set of input values. 
Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
Every state then encodes a set of output values to be produced each time the controller transitions to that state. 
As in all Moore machines, each state produces exactly one set of output values. 
Controllers synthesized from GR(1) specifications have the additional property that for each state, 
the set of input values that brings the controller into that state is the same, regardless of what the previous state was.
% there is exactly one set of input values that brings the controller into that state.

>>>>

Several tools for GR(1) synthesis are available. 
For example, RATSY %(Requirements Analysis Tool with Synthesis) 
\cite{bloem2010ratsy} has a focus on circuit design and synthesizes designs 
encoded in BLIF, %(Berkeley Logic Interchange Format),
Verilog, and HIF. %(HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP 
%(Linear Temporal Logic MissiOn Planning)  
\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
TuLiP 
%(Temporal Logic Planning) 
\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
Slugs 
%(SmalL bUt Complete GROne Synthesizer) 
\cite{Ehlers2016} is architected to allow users to tailor synthesis algorithms,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
 %% -- TODO: put more details about input types that Salty allows? About what optimizations it does?
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable software implementations in a variety of programming languages, 
including Python, C++, Java, and now SPARK. 
Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,''
tool implementation errors could still result in errors in synthesized controllers.
We are therefore interested in ``end-to-end'' verification of synthesized controllers.
Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK.

SPARK, which is based on the Ada programming language, is both a programming language 
with a specification language and an associated verification toolset \cite{hoang2015spark}.
SPARK translates code to the Why language for processing by the Why3 theorem prover, 
and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
For example, if one were to translate source code originally written in Ada to SPARK, 
one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
Formal verification with SPARK also requires supplying requirements at the level of the source code in the 
form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by indvidual subprograms.
There are different levels to which one can use SPARK to verify a program. 
Generally, the level of verification performed during development is incremental, going from lowest to highest. 
Verification at the lowest level, ``stone level,'' is achieved when code is accepted by SPARK, 
since the SPARK language has stricter legality rules than Ada. 
Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
flow analysis checks for errors such as reads of unintialized data or violations of user-specified data flow contracts. 
Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
e.g. division by zero and numeric overflow/underflow. 
Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
type invariants and subprogram preconditions and postconditions; 
however, at this level, the specifications only partially specify the desired behavior of the code.  
Verification at the next level, ``platinum level,'' consists verifying a complete set of specifications.
At the moment, the SPARK code generated by Salty proves at gold level.  
SPARK is able to prove the system initial and system transition property portions of the original GR(1) specification. 
Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
The platinum level would be reached if the liveness properties were expressed and proved, 
which we leave for future work.

@@

We have extended the open source tool Salty\footnote{https://github.com/GaloisInc/salty/} to produce software implementations of controllers synthesized from GR(1) specifications in SPARK. 
Given a GR(1) specification in Salty format, Salty does some preprocessing to sanity check, optimize, and translate the specification to Slugs format. 
Salty then calls Slugs to synthesize a controller design from the processed specification. 
Slugs provides the option of returning the controller design as a Moore machine expressed in text format, 
which is what Salty uses to create an executable software implementation of the controller. 
This Moore machine encodes all the states of the controller. %, including a single initial state. 
It also encodes the transition relation between states, i.e. the controller's next state given the current state 
and the next set of input values. 
Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
Every state then encodes a set of output values to be produced each time the controller transitions to that state. 
As in all Moore machines, each state produces exactly one set of output values. 
Controllers synthesized from GR(1) specifications have the additional property that for each state, 
the set of input values that brings the controller into that state is the same, regardless of what the previous state was.
% there is exactly one set of input values that brings the controller into that state.
